A decade of lost precision: Introducing a general framework for image feature matching without geometric constraints

### 2014-04-14 text/journal/paper.tex
Various improvements in terms of clarity

In the author's last paper we introduced \emph{Mirror Match} and 
\emph{Mirror-Match with Clustering}, two algorithms that we showed to 
perform better than the state of the art. The novel contributions of 
this paper consists of two new algorithms completing a framework of four 
related matching algorithms as well as the theoretical foundation which 
enables us to understand why \emph{Mirror Match} performs better than 
\emph{Ratio Match} in the first place.

@@

Given two images we can construct the proposal and baseline sets in six 
different ways as illustrated in Figure~\ref{fig:overview}. In the case 
of \emph{Ratio-Match} \cite{lowe2004sift} the \emph{proposal set} 
consists of all the feature points in the target image.  Similarly the 
\emph{baseline set} consists of all the feature points in the target 
image except the \emph{proposed} match. For \emph{Ratio-Match} this 
practically means that given a query feature, $f_q$, the \emph{proposed 
match} and the \emph{baseline match} are the two nearest neighbors in 
the set of feature points in the \emph{target image} and the ratio is 
decided uniquely based on the \emph{target image}. In turn 
\emph{Mirror-Match-Ext} as introduced by the authors in 
\cite{arnfred2013mirror}, works by constructing the \emph{proposal set} 
and \emph{baseline set} by combining features from both the \emph{query} 
and \emph{target image}. By similar compositions of the \emph{proposal} 
and \emph{baseline set} we derive four more algorithms as illustrated in 
Figure~\ref{fig:overview}.

@@

% TODO: include point about two of the algorithms being identical
%
\subsection{Generalized Uniqueness Ratio}
%

@@

Based on these assumptions we first prove that 
The above proof can easily be extended to show that \emph{Mirror-Match 
1} and \emph{Mirror-Match-Ext} returns the same correspondences too, and 
both will be referred to as \emph{Mirror Match} from here on.

For the case of \emph{Ratio-Match} and \emph{Ratio-Match-Ext} however we 
can show that \emph{Ratio-Match-Ext} is equal or better than 
\emph{Ratio-Match}.  Consider again the nearest neighbor $f_{nn}$ of a 
query feature $f_{q}$ and the two cases where we either have the $f_{nn} 
\in F_{query}$ or $f_{nn} \in F_{target}$. For $f_{nn} \in F_{target}$ 
the uniqueness ratio of \emph{Ratio-Match-Ext} is:
\begin{align*}
    r &= \text{r}(f_{q}, F_{proposed}, F_{baseline}) \\
        &= \text{r}(f_{q}, F_{query} \cup F_{target}, F_{baseline})\\
        &= \text{r}(f_{q}, F_{target}, F_{baseline})
\end{align*}

@@

Where as \emph{Ratio-Match-Ext} does not return the correspondence if 
the nearest neighbor of a query feature is in the \emph{query image}, 
\emph{Ratio-Match} is likely to do so, since the uniqueness ratio is 
found by dividing with the second nearest neighbor in the \emph{target 
image}. This shows us that for cases where a query feature has a nearest 
neighbor in the \emph{query image}, \emph{Ratio-Match-Ext} is superior 
to \emph{Ratio-Match}. Since query features with a nearest neighbor in 
the \emph{query image} are false correspondences per assumption \#1 and 
\#2, this proves that \emph{Ratio-Match-Ext} has superior 
\emph{Precision} to \emph{Ratio-Match} while maintaining equal 
\emph{Recall}.

Finally we show that \emph{Mirror Match} is equal or better than 
\emph{Ratio-Match-Ext}. Consider as before the nearest neighbor $f_{nn}$ 
of a query feature $f_{q}$. When $f_{nn}$ resides in the \emph{query 
image} both algorithms behave alike and discard the match.  However, 
consider the uniqueness ratio of \emph{Mirror Match} for the case where 
$f_{nn}$ resides in the \emph{target image}:
\begin{align*}
    r &= \text{r}(f_{q}, F_{proposed}, F_{baseline}) \\
        &= \text{r}(f_{q}, F_{proposed}, F_{query} \cup F_{target})\\
        &= max( \text{r}(f_{q}, F_{proposed}, F_{target}), 
    \text{r}(f_{q}, F_{proposed}, F_{query}) )
\end{align*}

For a true correspondence we have that the uniqueness ratio using 
$F_{query}$ as baseline is distributed similarly to the ratio when using 
$F_{target}$ by assumption \#3. In this case the algorithm performs like 
\emph{Ratio-Match-Ext}.  However for an untrue correspondence with a 
baseline match in the \emph{query image} which is closer than the
baseline match in the \emph{target image}, \emph{Mirror Match} will 
return a worse uniqueness ratio. This in turns means that \emph{Mirror 
Match} is equal or better than \emph{Ratio-Match-Ext} in terms of  
\emph{Precision} while maintaining equal \emph{Recall}.

For any given recall we have proven based on assumption \#1-3 that the 
precision of the algorithms in the framework compares as follows:
\begin{align*}
    \textit{Self-Match} &= \textit{Self-Match-Ext} \\
    \textit{Mirror-Match} &= \textit{Mirror-Match-Ext} \\
    \textit{Ratio-Match} &\leq \textit{Ratio-Match-Ext} \leq 
    \textit{Mirror Match}
\end{align*}

The difference between \emph{Self-Match} and \emph{Ratio-Match} is is 
the source of the \emph{baseline set}. For true correspondences we 
assume that they should perform similarly but as for the rest the 
performance of the two depends entirely on the images we are matching. 
When we match images where the target image might not overlap at all 
with the query image, using the query image and not the target image as 
a \emph{baseline set} seems like a reasonable choice given that the 
baseline set in this case would be closer to the feature matched and 
more strictly rule out untrue correspondences.

@@

According to the second assumption, feature descriptors behave such that
when matched, a true correspondence to a query feature will always be 
the nearest neighbor. For actual feature detectors and descriptors 
Mikolajczyk and Tuytelaars shows that this property depends strongly on 
the differences between \emph{query} and \emph{target} image 
\cite{mikolajczyk2005performance} \cite{tuytelaars2008local}. In the 
degenerate case where the closest match is not an actual correspondence, 
it is often the case that other matches are similarly close resulting in 
a uniqueness ratio near 1, effectively filtering out bad 
correspondences.  This means that with a lenient threshold close to 1, 
the guarantees about precision and recall equivalence in between the 
different ratio algorithms might no longer hold due to the lack of 
consistent descriptor behaviour. It is worth pointing out that if this 
assumption holds true, matching using only nearest neighbor is not 
necessarily optimal unless there is a complete scene overlap between
the images matched. For cases where there is little or no overlap, a 
nearest neighbor would indiscriminately return false matches even with 
well behaved feature descriptors.

@@

The last assumption states that the uniqueness ratio using either the 
\emph{query} or \emph{target image} as the baseline set would be similar
in distribution when we know that the feature point we are matching has 
a true correspondence. For an ideal feature descriptor where any true 
correspondence between two feature points will have a distance of $0$, 
the third assumption is obviously true for any cases where the second 
closest feature point has a true correspondence in the other image. In 
the general case it is likely that the baseline yields different 
uniqueness ratios depending on the features it contains, but for the 
case where the baseline match has a true correspondence in the other 
image we can verify the assumption by looking at the performance of 
\emph{Self-Match} compared with \emph{Ratio-Match}. To compare we plot 
the actual \emph{uniqueness ratios} measured from 84 3D objects in 
Figure~\ref{fig:ratio_hist}. For ratios lower than $0.7$ this is 
generally the case, but as we approach more lenient thresholds the 
performance of \emph{Mirror-Match} might no longer be equal or better 
than \emph{Ratio-Match-Ext} in terms of precision.
% Needs rewording

@@

The results can be seen in Figure~\ref{fig:all_objects}. In the figure 
we see the performance of the different matching methods in our proposed
framework for $12$ increasingly bigger rotations. The results are shown 
in a precision / recall plot to make it easy to compare performance in 
terms of precision at similar levels of recall.

@@

In the middle row featuring rotations of between $25$ and $40$ degrees 
we see the strongest improvement of \emph{Mirror-Match} compared to 
\emph{Ratio-Match} generally peaking at around
a 20 percentage-point increase in precision. Finally the the performance
gain decreases as we further increase the rotation to between $45$ and 
$60$. We suspect that this is partly because assumption \#2 that a 
descriptor is well behaved and match best with a descriptor of a true 
correspondence stops being valid when the images are transformed beyond 
a certain level of recognizability.

@@

\begin{table}[htb]
\caption{Complexity and running times tested on 15 image pairs on 3 
lighting conditions with average $n = 236.73$ and average $m = 247.38$ 
as tested on a Intel\textregistered\ Core\texttrademark\ i5-3550 CPU @ 
3.30~GHz with 8~GB memory.}
\label{table:running_times}
	\centering
%	\small
    \begin{tabular}{{l}{l}{l}}
    Algorithm & Complexity & Running Time\\
    \hline
    \noalign{\smallskip}
    %
    \emph{Self-Match} & $O(n\log(nm))$ & 118s  \\
    %\emph{Self-Match-Ext} & $O(n\log(n(n+m)))$ & 109s\\
    \emph{Ratio-Match} & $O(n\log(m))$ & 114s\\
    \emph{Ratio-Match-Ext} & $O(n\log(n+m))$ & 112s\\
    \emph{Mirror-Match} & $O(n\log(n(n+m)))$ & 116s\\
    %\emph{Mirror-Match-Ext} & $O(n\log(n+m))$ & 110s \\
    \hline
\end{tabular}
\end{table}

@@

\appendix
\section{Proof of equivalence between \emph{Self-Match} and 
\emph{Self-Match-Ext}}
In this appendix we prove the equivalence between \emph{Self-Match} and 
\emph{Self-Match-Ext}. To do this, we look at the nearest neighbor 
$f_{nn}$ of every query feature $f_{q}$ and consider the two cases of 
$f_{nn} \in F_{query}$ and $f_{nn} \in
F_{target}$. For $f_{nn} \in F_{target}$ the uniqueness ratio of 
\emph{Self-Match-Ext} is:
\begin{align*}
    r &= \text{r}(f_{q}, F_{proposed}, F_{baseline}) \\
        &= \text{r}(f_{q}, F_{query} \cup F_{target}, F_{baseline})\\
        &= \text{r}(f_{q}, F_{target}, F_{baseline})
\end{align*}

@@

\section{Proof of equivalence between \emph{Mirror-Match} and 
\emph{Mirror-Match-Ext}}

In this appendix we prove the equivalence between \emph{Mirror-Match} 
and \emph{Mirror-Match-Ext}. This proof is almost identical to the 
equivalence proof between \emph{Self-Match} and \emph{Self-Match-Ext}. 
To prove this, we look at the nearest neighbor $f_{nn}$ of every query 
feature $f_{q}$ and consider the two cases of $f_{nn} \in F_{query}$ and 
$f_{nn} \in
F_{target}$. For $f_{nn} \in F_{target}$ the uniqueness ratio of 
\emph{Mirror-Match-Ext} is:
\begin{align*}
    r &= \text{r}(f_{q}, F_{proposed}, F_{baseline}) \\
        &= \text{r}(f_{q}, F_{query} \cup F_{target}, F_{baseline})\\
        &= \text{r}(f_{q}, F_{target}, F_{baseline})
\end{align*}

Since $F_{proposed} = F_{target}$ for \emph{Mirror-Match} this shows that
when the nearest neighbor is found in the target image, the two 
algorithms behave identically. For $f_{nn} \in F_{query}$ we know that 
the best match for the query feature is in the query image itself. Since 
neither algorithm returns a correspondence within the same image, they 
also behave identically for this case. This proves that for any set of 
\emph{query features} \emph{Mirror-Match} returns the same matches as 
\emph{Mirror-Match-Ext}.

>>>>

In the author's last paper we introduced \emph{Mirror-Match} and 
\emph{Mirror-Match with Clustering}, two algorithms that we showed to 
perform better than the state of the art. The novel contributions of 
this paper consists of two new algorithms completing a framework of four 
related matching algorithms as well as the theoretical foundation which 
enables us to understand why \emph{Mirror-Match} performs better than 
\emph{Ratio Match} in the first place.

@@

Given two images we can construct the proposal and baseline sets in six 
different ways as illustrated in Figure~\ref{fig:overview}. In the case 
of \emph{Ratio-Match} \cite{lowe2004sift} the \emph{proposal set} 
consists of all the feature points in the target image.  Similarly the 
\emph{baseline set} consists of all the feature points in the target 
image except the \emph{proposed} match. For \emph{Ratio-Match} this 
practically means that given a query feature, $f_q$, the \emph{proposed 
match} and the \emph{baseline match} are the two nearest neighbors in 
the set of feature points in the \emph{target image} and the ratio is 
decided uniquely based on the \emph{target image}. In turn 
\emph{Mirror-Match} as introduced by the authors in 
\cite{arnfred2013mirror}, works by constructing the \emph{proposal set} 
and \emph{baseline set} by combining features from both the \emph{query} 
and \emph{target image}. By similar variations of the \emph{proposal} 
and \emph{baseline set} we derive four more algorithms as illustrated in 
Figure~\ref{fig:overview}, however we will not discuss 
\emph{Self-Match-Ext} and \emph{Both-Match} further in this paper since 
we show them to be equivalent with \emph{Self-Match} and 
\emph{Mirror-Match} respectively in Appendix~\ref{A:self} and 
\ref{A:mirror}.

@@

%
\subsection{Generalized Uniqueness Ratio}
%

@@

Based on these assumptions we prove that \emph{Ratio-Match-Ext} is equal 
or better than \emph{Ratio-Match} in terms both precision and recall.  
Consider the nearest neighbor $f_{nn}$ of a query feature $f_{q}$ and 
the two cases where we either have the $f_{nn} \in F_{query}$ or $f_{nn} 
\in F_{target}$. For $f_{nn} \in F_{target}$ the uniqueness ratio of 
\emph{Ratio-Match-Ext} is:
\begin{align*}
    r &= \text{r}(f_{q}, F_{proposed}, F_{baseline}) \\
        &= \text{r}(f_{q}, F_{query} \cup F_{target}, F_{baseline})\\
        &= \text{r}(f_{q}, F_{target}, F_{baseline})
\end{align*}

@@

That is, \emph{Ratio-Match} calculates the ratio based on the two 
nearest correspondences in the \emph{Target Image}. 
\emph{Ratio-Match-Ext} on the other hand does not return any 
correspondence because the nearest neighbor is in the \emph{Query 
Image}. Since query features with a nearest neighbor in the \emph{query 
image} are false correspondences per assumption \#1 and \#2, this proves 
that \emph{Ratio-Match-Ext} has superior \emph{Precision} to 
\emph{Ratio-Match} while maintaining equal \emph{Recall}.

Next we show that \emph{Mirror-Match} is equal or better than 
\emph{Ratio-Match-Ext} in terms of precision and recall. Consider as 
before the nearest neighbor $f_{nn}$ of a query feature $f_{q}$. When 
$f_{nn}$ resides in the \emph{query image} both algorithms behave alike 
and discard the match.  However, consider the uniqueness ratio of 
\emph{Mirror-Match} for the case where $f_{nn}$ resides in the 
\emph{target image}:
\begin{align*}
    r &= \text{r}(f_{q}, F_{proposed}, F_{baseline}) \\
        &= \text{r}(f_{q}, F_{proposed}, F_{query} \cup F_{target})\\
        &= max( \text{r}(f_{q}, F_{proposed}, F_{target}), 
    \text{r}(f_{q}, F_{proposed}, F_{query}) )
\end{align*}

For a true correspondence we have that the uniqueness ratio using 
$F_{query}$ as baseline is distributed similarly to the ratio when using 
$F_{target}$ by assumption \#3. In this case the algorithm performs like 
\emph{Ratio-Match-Ext}.  However for an untrue correspondence with a 
baseline match in the query image which is closer than the
baseline match in the target image, \emph{Mirror-Match} will return a 
worse uniqueness ratio. This in turns means that \emph{Mirror-Match} is 
equal or better than \emph{Ratio-Match-Ext} in terms of  
\emph{Precision} while maintaining equal \emph{Recall}.

Using a similar procedure we prove that \emph{Self-Match} is equal to
\emph{Self-Match-Ext} and \emph{Both-Match} is equal to \emph{Mirror
Match} in Appendix~\ref{A:self} and \ref{A:mirror} respectively.

For any given recall we have proven based on assumption \#1-3 that the 
precision of the algorithms in the framework compares as follows:
\begin{align*}
    \textit{Ratio-Match} &\leq \textit{Ratio-Match-Ext} \leq 
    \textit{Mirror-Match}
\end{align*}

The difference between \emph{Self-Match} and \emph{Ratio-Match} is
the source of the \emph{baseline set}. For true correspondences we 
assume that they should perform similarly but as for the rest the 
performance of the two depends entirely on the images we are matching. 
When we match images where the target image might not overlap at all 
with the query image, using the query image and not the target image as 
a \emph{baseline set} seems like a reasonable choice given that the 
baseline set in this case would be closer to the feature matched and 
more strictly rule out untrue correspondences.

@@

According to the second assumption, feature descriptors behave such that
when matched, a true correspondence to a query feature will always be 
the nearest neighbor. For actual feature detectors and descriptors 
Mikolajczyk and Tuytelaars shows that this property depends strongly on 
the differences between query and target image 
\cite{mikolajczyk2005performance} \cite{tuytelaars2008local}. In the 
degenerate case where the closest match is not an actual correspondence, 
it is often the case that other matches are similarly close resulting in 
a uniqueness ratio near 1, effectively filtering out bad 
correspondences.  This means that with a lenient threshold close to 1, 
the guarantees about precision and recall equivalence in between the 
different ratio algorithms might no longer hold due to the lack of 
consistent descriptor behaviour. It is worth pointing out that if this 
assumption holds true, matching using only nearest neighbor is not 
necessarily optimal unless there is a complete scene overlap between
the images matched. For cases where there is little or no overlap, a 
nearest neighbor would indiscriminately return false matches even with 
well behaved feature descriptors.

@@

The last assumption states that the uniqueness ratio using either the 
\emph{query} or target image as the baseline set would be similar
in distribution when we know that the feature point we are matching has 
a true correspondence. For an ideal feature descriptor where any true 
correspondence between two feature points will have a distance of $0$, 
the third assumption is obviously true for any cases where the second 
closest feature point has a true correspondence in the other image. For 
non-ideal feature descriptors where the baseline match has a true 
correspondence in the other image we can verify the assumption by 
looking at the \emph{uniqueness ratio} for a true correspondences when 
it is based on the query image compared to when it is based on the 
target image. To compare we plot the actual \emph{uniqueness ratios} 
measured from $84$ 3D objects in Figure~\ref{fig:ratio_hist}. For ratios 
lower than $0.7$ this is generally the case, but as we approach more 
lenient thresholds the performance of \emph{Mirror-Match} might no 
longer be equal or better than \emph{Ratio-Match-Ext} in terms of 
precision.
% Needs rewording

@@

The results can be seen in Figure~\ref{fig:all_objects}. In the figure
we see the performance of the different matching methods in our proposed
framework for $12$ increasingly bigger rotations. The results are shown
in a precision / recall plot to make it easy to compare performance in
terms of precision at similar levels of recall. For the application of
feature matching it is important to keep in mind that for many
applications, such as image stitching, object and scene recognition and
near duplicate detection, precision is much more important than recall,
since a handfull of highly reliable correspondences tells us much more
about the similarity of image components than a high number of imprecise
matches. For applications where we intent to combine one of the
algorithms with a geometric method, we would ideally like for it to
produce results with a high recall that can then be further filtered by
a geometric method.

@@

In the middle row featuring rotations of between $25$ and $40$ degrees 
we see the strongest improvement of \emph{Mirror-Match} compared to 
\emph{Ratio-Match} generally peaking at around
a $20$ percentage-point increase in precision. Finally the the performance
gain decreases as we further increase the rotation to between $45$ and 
$60$. We suspect that this is partly because assumption \#2 that a 
descriptor is well behaved and match best with a descriptor of a true 
correspondence stops being valid when the images are transformed beyond 
a certain level of recognizability.

@@

\begin{table}[htb]
\caption{Complexity and running times tested on 45 image pairs with average $n = 237$ and average $m = 247$ 
as tested on a Intel\textregistered\ Core\texttrademark\ i5-3550 CPU @ 
3.30~GHz with 8~GB memory.}
\label{table:running_times}
	\centering
%	\small
    \begin{tabular}{{l}{l}{l}}
    Algorithm & Complexity & Running Time\\
    \hline
    \noalign{\smallskip}
    %
    \emph{Self-Match} & $O(n\log(nm))$ & 118s  \\
    %\emph{Self-Match-Ext} & $O(n\log(n(n+m)))$ & 109s\\
    \emph{Ratio-Match} & $O(n\log(m))$ & 114s\\
    \emph{Ratio-Match-Ext} & $O(n\log(n+m))$ & 112s\\
    %\emph{Both-Match} & $O(n\log(n(n+m)))$ & 116s\\
    \emph{Mirror-Match} & $O(n\log(n+m))$ & 110s \\
    \hline
\end{tabular}
\end{table}

@@

\appendix

\subsection{Proof of equivalence of Self-Match and Self-Match-Ext}
\label{A:self}

In this appendix we prove the equivalence between \emph{Self-Match} and 
\emph{Self-Match-Ext}. To do this, we look at the nearest neighbor 
$f_{nn}$ of every query feature $f_{q}$ and consider the two cases of 
$f_{nn} \in F_{query}$ and $f_{nn} \in
F_{target}$. For $f_{nn} \in F_{target}$ the uniqueness ratio of 
\emph{Self-Match-Ext} is:
\begin{align*}
    r &= \text{r}(f_{q}, F_{proposed}, F_{baseline}) \\
        &= \text{r}(f_{q}, F_{query} \cup F_{target}, F_{baseline})\\
        &= \text{r}(f_{q}, F_{target}, F_{baseline})
\end{align*}

@@

\subsection{Proof of equivalence of Both-Match and Mirror-Match}
\label{A:mirror}

In this appendix we prove the equivalence between \emph{Both-Match} and 
\emph{Mirror-Match}. This proof is almost identical to the equivalence 
proof between \emph{Self-Match} and \emph{Self-Match-Ext}. To prove 
this, we look at the nearest neighbor $f_{nn}$ of every query feature 
$f_{q}$ and consider the two cases of $f_{nn} \in F_{query}$ and $f_{nn} 
\in
F_{target}$. For $f_{nn} \in F_{target}$ the uniqueness ratio of 
\emph{Mirror-Match} is:
\begin{align*}
    r &= \text{r}(f_{q}, F_{proposed}, F_{baseline}) \\
        &= \text{r}(f_{q}, F_{query} \cup F_{target}, F_{baseline})\\
        &= \text{r}(f_{q}, F_{target}, F_{baseline})
\end{align*}

Since $F_{proposed} = F_{target}$ for \emph{Both-Match} this shows that
when the nearest neighbor is found in the target image, the two 
algorithms behave identically. For $f_{nn} \in F_{query}$ we know that 
the best match for the query feature is in the query image itself. Since 
neither algorithm returns a correspondence within the same image, they 
also behave identically for this case. This proves that for any set of 
\emph{query features} \emph{Both-Match} returns the same matches as 
\emph{Mirror-Match}.
